\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-04 chapters/gaze_independence.tex
Update bttda formulation

\todo{talk about covert shifts in spatial attention alpha}

\section{Eye motor impairment in individuals with \acs{sspi}}%
\label{sec:gaze-dependence}

\todo{define eye motor issues}

\todo{Algemeen: in begin uitleggen dat de ziektes er niet veel toe doen en dat
het gaat over verlamde personen, verlies van spierkracht en coördinatie en dus
gehinded in communicatie, helemaal in begin van chapter 1 duidelijk maken dat
het om deze patiënten gaat}

One of the goals of brain-computer interfacing is establishing a communication
channel that does not rely on speech or muscular activity which in turn can
provide solutions to individuals with \ac{sspi}.
In the strictest interpretation, this then means that an interface should not
rely on the control of eye muscles used to redirect the gaze or for blinking.
It is exactly this potential of \acp{bci} that makes them suitable as assistive
communication devices.
\todo{to the CNS (stroke, spinal chord injury, tbi) or neurodegenerative diseases}

Unfortunately, many individuals in the \ac{bci} target population suffer to
some extent from eye motor disability, requiring \ac{bci}s adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor problems, which can range from minor (nystagmus
\footnote{%
Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).
}, other eye tremors
\footnote{%
These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.
}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
\footnote{%
Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.
}, involuntary movements, impaired pursuit, \ldots), and even complete
ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only has an effect on vision and coordination but also
on their ability to operate a visual \ac{bci}.

\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
    & \bfseries \acs{als} & \bfseries \acs{ms}   & \bfseries Stroke &\bfseries\acs{dmd} &\bfseries \acs{sma} &\bfseries \acs{lis} \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
  \caption[Incidence of eye motor impairment in selected \ac{bci} target
    patient populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} target
      patient populations. (\acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{lis}: \acl{lis}).
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}
\todo{include FA}
\todo{cite Abnormal Oculomotor Functions in Amyotrophic Lateral Sclerosis 10.3988/jcn.2018.14.4.464}
\todo{cite Neuro-Ophthalmologic Manifestations of Amyotrophic Lateral Sclerosis}

\todo{
14] Fazzi E, Signorini SG, Piana RL, et al. Neuro-ophthalmo-
logical disorders in cerebral palsy: ophthalmological, oculo-
motor, and visual aspects. Dev Med Child Neurol. 2012;54:
730–736.
[15] Graber M, Challe G, Alexandre MF, et al. Evaluation of the
visual function of patients with locked-in syndrome: report
of 13 cases. J Fr Ophtalmol. 2016;39:437–440.
[16] Moss HE, McCluskey L, Elman L, et al. Cross-sectional evalu-
ation of clinical neuro-ophthalmic abnormalities in an
amyotrophic lateral sclerosis population. J Neurol Sci. 2012;
314:97–101.
}

Among the most affected patients are stroke patients~\cite{Pollock2011}, mostly
due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
\todo{summarize and cite Rowe2019}
These patients might suffer from severe or complete eye motor impairment
from the onset of their condition.
In \ac{als}, a progressive disease, oculomotor issues are also fairly common.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, \textcite{Guo2022} show that minor issues are still
fairly common,
especially for patients with bulbar onset \ac{als}.
Furthermore, \textcite{Hayashi1991} show that, as these patients progress past
the point of independent breathing, symptoms will eventually also involve
eye muscle paralysis.
Various forms of eye movement abnormalities also occur often and are especially well studied in
\ac{ms}~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020}.
These abnormalities can be minor or severe, seldom progressing to complete
paralysis.
However, \ac{ms} often comes with vision loss, further complicating interaction
with visual \acp{bci}.
\todo{FA}
Patients suffering from other neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities are
mostly preserved.

\newcommand\fnlis{\footnote{
Multiple definitions of \ac{lis} are encountered in
\ac{bci} and neurological literature.
Some definitions include only tetraplegic patients without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include stroke or \ac{tbi} patients with damage to
specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005},
it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late stage ALS.
In clinical reality, every tetraplegic patient has their own unique set of
preserved capabilities and from a solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
}}

@@

The most severely impaired of these patients toghether constitute the \ac{lis}
patient group, which forms one of the main \ac{bci} target populations.
In this work, we define \ac{lis} patients as those in a situation of (near)
complete paralysis and difficulties or the inability to communicate independently,
even with assistive technology.
This corresponds to classes one and two defined by
\textcite{Wolpaw2006}\fnwolpawcrit.
Additionally, some degree of severe or complete eye motor impairment is usually
necessary to qualify as locked-in\fnlis.
Case study reports~\cite{Patterson1986} show that the population of locked-in
patients with complete eye motor paralysis is very small.
Hence, it would be interesting to focus \ac{bci} development efforts on the
larger population of individuals with \ac{sspgi} that currently slip through
the cracks of the assistive technology offer.
These are individuals whose severe eye motor impairment prevents them from
using eye tracking based solutions.
They currently use their remaining motor control to
communicate by indicating symbols on a letterboard with great effort,
or signal with upwards eye movements or blinks to confirm prompted letters.
They require a caregiver or relative to interpret their signals, and crave
the ability to communicate independently which is crucial to retaining an
acceptable quality of life.

@@

@@

\section{State-of-the-art}
\label{sec:gaze-independence}
\todohsm{Address more systematically.}
\todo{in a paragraph describe each approach and its up and downsides}
\todo{rewrite}

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and
somatosensory\todo{cite some}
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
While auditory and somatosensory stimulation on its own might yield poor
\ac{itr} \textcite{Yin2016} that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94bits/min.
In general, their approach is interesting since they follow a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

\todo{insert table from riccio in text as numbers, not as table}

\subsection{Gaze-independent visual stimulation}
This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm in particular is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the field of view of the user.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}\todo{cite},
where targets or small sets of targets are shown consecutively.
The Hex-o-Spell, Geospell and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interface
have historically shown not to perform well in the presence of gaze impairment.
However, we argue that, spatially organized interfaces in general have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achive an \ac{itr} to up to
30bits/min.
With the right set of adaptations, spatial attention could potentionally also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces they also provide an extra channel that can be used to transfer
information, i.e.\ the location of the stimulus.

As an alternative to spatial attention, non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
The \ac{rsvp} speller is already an example of this, relying on the user to
attend to attend the appearance of the intended character to type.
Visual stimulation paradigms relying on alternative types of attention
can modulate specific \ac{erp} components
that are more sensitive to stimulation in the visual
periphery~\cite{Schaeff2012,Xu2022}.\todo{discuss Xu2022}
Nevertheless, these methods can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
Furthermore, the previously mentioned systems typically were only tested in
settings where the user was required to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.

@@

The general goal of this PhD is to tackle the gaze-dependence in problem in the
visual oddball \ac{bci}.
\todo{refer to ITRs from previous chapter}
Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

This leads us to adopt a specific approach: improve the performance of a
spatially organized visual oddball \ac{erp}-based brain
computer interface by using a suited decoding strategy.
Hence, our working hypothesis is as follows:
\todo{rewrite, this is not what we tested in the patient paper after all}
\begin{quote}
	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
		can achieve a similar transfer rate in patient
		populations with eye motor deficits than other gaze-independent alternatives
    described in literature.}
\end{quote}

To achieve our goal, we innovate on decoder development and interface design,
and we collect data from healthy control subjects, which allow us to test our decoders.
Finally, the findings of these experiments will be leveraged to develop a \ac{bci}
for patient use, which will be tested in a case-based patient study in which
both performance and user comfort was be evaluated.\todo{was this actually
done?}

This specific hypothesis is embedded within a broader goal of enabling
effective communication for eye-motor impaired \ac{lis} patients.
Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities, by leveraging a non-invasive, high-\ac{itr}
\ac{bci} on the one hand, and on the other hand improving \ac{erp} decoding performance
in general.

@@

\subsection{Interface design}
\todohsm{rename data collection. start with decoder design, second section data collection with experiments}

\todo{mention, free, split, frenzel}
\todo{clear definitions, table from paper}

First, we designed a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
decoders and on our own proposed decoders.
Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
\textcite{Treder2010}, we present six flashing circular targets
(without letters or symbols) to the participant while the EEG and
electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
\todo{chosen because gaze-independent optimized and Peters2020}

@@

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
      The user is free to direct their gaze as they deem most comfortable
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
  }
  \label{fig:gaze-vsa}
\end{figure}
In the overt case, users gaze at the cued
target they are also mentally attending; in the covert case, users
gaze at the center of the screen while mentally attending to the cued target.
Finally, we introduce a VSA condition that is understudied in the context of
gaze-independent \ac{bci} development: split VSA.
In split VSA, the user mentally focuses on one cued target while gazing at
another (the distractor).
\todomvh{+ "free" condition in Fig. 21.}

The option to dissociate gaze and visuospatial attention will allow us to
investigate the effect of gaze control on \ac{bci} performance.

\subsection{Decoders}

During this PhD, we explored different lines in decoding strategies,
trying to tackle several problems that arise from gaze-independence, such as
the lack or decrease in amplitude of specific \ac{erp} components, and the increased
non-stationarity of the signal.
As mentioned, state-of-the-art decoders have poor performance in covert attention
settings.
The general goal is thus to design a machine learning classifier that represents
the \ac{erp} signal in a specific way, such that it becomes more robust to the problems
occurring in covert attention conditions.

\subsubsection{Regularized spatiotemporal beamforming}
Due to the decreased amplitude of the N1 and P3 components in covert attention
settings~\cite{Treder2010}, the \ac{snr} of
the \ac{erp} is lower than in overt attention settings.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance, might be by increasing overall \ac{erp}
decoding performance. A more accurate classifier could yield relative
improvements in those settings where performance is not yet near the achievable
maximum.\todomvh{?? how to find out what is the maximum?}

>>>>

\section{Eye motor impairment in individuals with \acl{sspi}}%
\label{sec:gaze-dependence}

One of the goals of brain-computer interfacing is establishing a communication
channel that does not rely on speech or muscular activity which in turn can
provide solutions to individuals with \ac{sspi}.
In the strictest interpretation, this then means that an interface should not
rely on the control of eye muscles used to redirect the gaze or for blinking.
It is exactly this potential of \acp{bci} that makes them suitable as assistive
communication devices.

\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups unfortunately also suffer to
some extent from eye motor impairment, requiring \ac{bci}s adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus
\footnote{%
Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).
}, other eye tremors
\footnote{%
These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.
}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
\footnote{%
Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.
}, involuntary movements, impaired pursuit, \ldots), and even complete
ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only has an effect on vision and coordination but also
on their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

\begin{table}
  %\sffamily
  %\footnotesize
	%\centering
  \input{chapters/gaze_independence_incidence.tex}
  \caption[Incidence of eye motor impairment in selected \acs{bci} target
    patient populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} target
      patient populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}

Among the most  severely affected patients are stroke
patients~\cite{Pollock2011,Rowe2019}, mostly
due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
These patients might suffer from severe or complete eye motor impairment
from the onset of their condition.
Individuals recovering from stroke with severe eye motor impairment are
Furthermore, those with
Case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
with \ac{lis} due to stroke, the group with a complete lack of eye motor
control is very small.

In \ac{als}, a progressive disease affecting motor neurons, oculomotor issues
are also fairly common.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, show that minor issues are still
fairly common, especially for patients with bulbar onset
\ac{als}~\cite{Kang2018, Guo2022,Moss2012}.
Furthermore, \textcite{Hayashi1991} show that, as these patients progress past
the point of independent breathing, symptoms will eventually also involve
eye muscle paralysis.

Various forms of eye movement abnormalities also occur often in
\ac{ms}.
\ac{ms} is a neurodegenerative disease involving demyelination of nerves.
Eye motor abnormalities are especially well
studied~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020} and
often used as diagnostical tool.
These abnormalities can be minor or severe, seldom progressing to complete
paralysis.
However, \ac{ms} often comes with vision loss, further complicating interaction
with visual \acp{bci}.

\ac{fa} is a neurodegenerative disease involving loss of coordinated movement.
This almost always affects eye
movements~\cite{Fahey2008,Hocking2010,Furman1983}, with various forms of
involuntary movements and troubles pursuing or fixating targets.
They also gradually have more troubles speaking, but often retain some muscular
control.

Some other groups that are heavily affected are \ac{cp}~\cite{Fazzi2012} and
\ac{fa}~\cite{Fahey2008,Hocking2010,Furman1983}.
Individuals with some neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities are
mostly preserved.

\newcommand\fnlis{\footnote{
Multiple definitions of \ac{lis} are encountered in
\ac{bci} and neurological literature.
Some definitions include only tetraplegic patients without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include stroke or \ac{tbi} patients with damage to
specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005},
it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late stage ALS.
}}

@@

We believe it is in general not opportune to delve too deep in the underlying
ophthalmological and neurological pathophysiology for each of the specific
conditions.
In clinical reality, every patient different symptoms, which lead to their own
unique set of preserved capabilities.
From a solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients we will refer to these groups as individuals
with \ac{sspi} and, if eye motor skills are affected,  \ac{sspgi}, in line with
the terminology used by~\textcite{FrieDOken2020} (see
Figure~\ref{fig:gaze_independence/venn}).

@@

The most severely impaired of these patients constitute the \ac{lis}
patient group, which forms one of the main \ac{bci} interest groups.
In this work, we use the term \ac{lis} for a situation of (near)
complete paralysis and difficulties or the inability to communicate independently,
even with assistive technology.
This corresponds to classes one and two defined by
\textcite{Wolpaw2006}\fnwolpawcrit.
\todo{[15] Graber M, Challe G, Alexandre MF, et al. Evaluation of the
visual function of patients with locked-in syndrome: report
of 13 cases. J Fr Ophtalmol. 2016;39:437–440.}
Hence, some degree of severe or complete eye motor impairment is usually
necessary to qualify as locked-in\fnlis.

In general, the group of locked-in individuals with complete loss of eye motor
function is very small.
Hence, it would be interesting to focus \ac{bci} development efforts on the
larger population of individuals with \ac{sspgi}  that currently slip through
the cracks of the assistive technology offer.
These are individuals whose severe eye motor impairment prevents them from
using eye tracking based solutions.
They currently use their remaining motor control to
communicate by indicating symbols on a letterboard with great effort,
or signal with upwards eye movements or blinks to confirm prompted letters.
They require a caregiver or relative to interpret their signals, and crave
the ability to communicate independently which is crucial to retaining an
acceptable quality of life.

@@

\section{State-of-the-art}
\label{sec:gaze-independence}

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010}
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
While auditory and somatosensory stimulation on its own might yield poor
\ac{itr} \textcite{Yin2016} that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94bits/min.
In general, their approach is interesting since they follow a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

\subsection{Gaze-independent visual stimulation}
This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm in particular is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the field of view of the user.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp},
where targets or small sets of targets are shown consecutively.
The Hex-o-Spell, Geospell and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interface
have historically shown not to perform well in the presence of gaze impairment.
However, we argue that, spatially organized interfaces in general have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achive an \ac{itr} to up to
30bits/min.
With the right set of adaptations, spatial attention could potentionally also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces they also provide an extra channel that can be used to transfer
information, i.e.\ the location of the stimulus.

As an alternative to spatial attention, non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
The \ac{rsvp} speller is already an example of this, relying on the user to
attend to attend the appearance of the intended character to type.
Visual stimulation paradigms relying on alternative types of attention
can modulate specific extra \ac{erp} components that either improve performance
because the embed extra information in the brain signal~\cite{Xu2022}
or because they are more sensitive to stimulation in the visual
periphery~\cite{Schaeff2012}.
Nevertheless, solely relying on alternative types of attenion can suffer too
from reduced information transfer rates~\cite{Chennu2013}.
Furthermore, the previously mentioned systems typically were only tested in
settings where the user was required to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.

@@

The general goal of this PhD is to tackle the gaze-dependence in problem in the
visual oddball \ac{bci}.
Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

This leads us to adopt a specific approach: \emph{improve the performance of a
spatially organized visual oddball \ac{erp}-based brain
computer interface by using a suited decoding strategy}.
%Hence, our working hypothesis is as follows:
%\todo{rewrite, this is not what we tested in the patient paper after all}
%\begin{quote}
%	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
%		can achieve a similar transfer rate in patient
%		populations with eye motor deficits than other gaze-independent alternatives
%    described in literature.}
%\end{quote}

To achieve our goal, we chose a suitable interface and innovate on decoder
development.
We collect data from healthy control subjects, which allow us to test our decoders.
Finally, the findings of these experiments will be verified in experiments with
individuals with \ac{sspgi}, including individuals with \ac{lis} evaluating
both performance and comfort.

This specific target is embedded within a broader goal of enabling
effective communication for eye-motor impaired \ac{lis} patients.
Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities.
We believe this can be done with a non-invasive, spatially
organized visual \acp{bci} with high-\ac{itr}.
Finally, or efforts also aim to improve hand improving \ac{erp} decoding performance
in general, which will also contribute to their effectiveness in gaze-dependent
or gaze-independent settings.

@@

\subsection{Decoder design}

During this PhD, we explored different lines in decoding strategies,
trying to tackle several problems, be it general problems in \ac{erp} decoding
or those that arise specifically from gaze-independence.
Examples of this are the lack or decrease in amplitude of specific \ac{erp} components, and the increased
non-stationarity of the signal.
As mentioned, state-of-the-art decoders have poor performance in covert attention
settings.
The general goal is thus to design a machine learning classifier that represents
the \ac{erp} signal in a specific way, such that it becomes more robust to the problems
occurring in covert attention conditions.

\subsubsection{Regularized spatiotemporal beamforming}
Due to the decreased amplitude of the N1 and P3 components in covert attention
settings~\cite{Treder2010}, the \ac{snr} of
the \ac{erp} is lower than in overt attention settings.
Therefore, a straightforward way to reach satisfactory
gaze-independent decoding performance, might be by increasing overall \ac{erp}
decoding performance.

Therefore, we have improved upon an in-house developed, state-of-the art \ac{erp}
decoder, the spatiotemporal beamformer~\cite{Wittevrongel2016}, by reformulating
this classifier as a linear discrimination problem and
imposing regularizing constraints by structuring the noise covariance matrix
(STBF-struct).
Furthermore, these regularizing constraints impose temporal stationarity on
the background noise, yielding insights for our next
efforts to cope with the non-stationarity of the P3 signal component.
This approach has been published by~\textcite{VanDenKerchove2022} and will be
described in Chapter~\ref{sec:stbf-struct}.

\subsubsection{Tensor discriminant analysis}
\todo{fill in}

\subsubsection{Classifier-based Latency Estimation with Woody iterations}
The previous two methods were aimed at improving decoding performance, but
did not (yet) yield results specifically in gaze-independent decoding.
The following approach is targeted towards improving decoding performance in
gaze-independent settings by accounting for a known property of the covert
attention \ac{erp} response.

P3 latency generally falls between 350ms and 600ms~\cite{Luck2014}, but this
value is heavily dependent on the subject and the task, and can vary from trial
to trial~\cite{Ouyang2017}.
The work of \cite{Arico2014} illustrates that the variation in single-trial P3
latencies is important in gaze-independent decoding and has been hampering covert
VSA decoding performance.
We reprised their hypothesis stating that jitter compensation through latency
estimation and alignment improves
covert \ac{vsa} performance and extend it by developing a decoder.

Existing latency estimation methods are either not applicable to the
classification problem of labeling unseen data, or are not robust enough to
deal with the low \ac{snr} of the \ac{erp}.
\Ac{cble}~\cite{Mowla2017} is a technique that can leverage
\ac{erp} latency estimation in a decoding setting, but our results show that it
yields no improvement in gaze-independent settings.
We improved upon this technique and extended it to a probabilistic, iterative
method named \ac{wcble}.
This approach has been published by~\cite{VanDenKerchove2024} and is presented
and evaluated on simulated data in Chapter~\ref{sec:wcble}.
We have also collected datasets to evaluate this approach on real \ac{erp} data
in gaze-independent settings.

\subsection{Data collection}

\subsubsection{The CVSA-ERP dataset}

In a first series of experiments, we recorded data with this interface from healthy participants.
The goal of these experiments is to benchmark
gaze-independent \ac{erp} decoding algorithms.
We denote this dataset as the Covert Visuospatial Attention \ac{erp} dataset
(CVSA-\ac{erp})
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
To simulate the dissociation between the eye gaze and the
intended target (\ac{vsa}) which could occur in individuals with gaze
impairment, healthy participants are cued to operate the \ac{bci} in specific
\ac{vsa} conditions.

\todo{mention, free, split, frenzel}

We chose a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the \ac{erp}, when using existing state-of-the-art
decoders and on our own proposed decoders.
Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
\textcite{Treder2010}, we present six flashing circular targets
(without letters or symbols) to the participant while the EEG and
electro-oculogram (EOG) are recorded, as well as eye gaze using eye tracking.
The Hex-o-Spell was chosen since it is optimiazed for gaze-independent
performance and because it has already been tested in individuals with
\ac{sspi} by~\textcite{Severens2014}.

@@

  \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_split.eps}
    \caption[Split \ac{vsa}]{%
      \emph{Split} \ac{vsa}.
      \Ac{vsa} is directed towards a target, while the gaze rests on another.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
   \begin{subfigure}[t]{.45\textwidth}
    \includegraphics[width=\textwidth]{figures/gaze_independence/attention_free.eps}
    \caption[Free \ac{vsa}]{%
      \emph{Free} \ac{vsa}.
      Gaze and \ac{vsa} coincide on a target.
      The user is free to direct their gaze as they deem most comfortable
      All previous \ac{vsa} conditions are possible.
    }
    \label{fig:gaze/vsa/overt}
  \end{subfigure}\hfill%
  \caption[\Ac{vsa} conditions]{%
    \Ac{vsa} conditions defined in our hexagonal spatial \ac{erp} paradigm interface.
  }
  \label{fig:gaze-vsa}
\end{figure}

In the overt case, users gaze at the cued
target they are also mentally attending; in the covert case, users
gaze at the center of the screen while mentally attending to the cued target.
Finally, we introduce a VSA condition that is understudied in the context of
gaze-independent \ac{bci} development: split VSA.
In split VSA, the user mentally focuses on one cued target while gazing at
another (the distractor).
This last option has only scarcely be studied, but completes
the option s to dissociate gaze and visuospatial attention, which allow us to
investigate the effect of (the lack of) gaze control on \ac{bci} performance.
This dataset and performance of our proposed~\ac{wcble} decoder has also been
published by~\textcite{VanDenKerchove2024} and will be presented in
Chapter~\ref{sec:covert-align}.
Performance is also evaluated on a publicly available dataset.

\subsubsection{Case studies with gaze-impaired individuals}
\todo{fill in}
\todo{mention multi-center patient study case studies in cooperation
After data collection from healthy participants and decoder development, we
have carry out a patient study using a similar interface.
The goal of the patient study is to evaluate the influence of eye motor
deficits in patient populations on existing state-of-the-art decoders and on
our proposed decoding methods.
}
The results of this study are presented in chapter~\ref{sec:patients}.
